Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-03-03 docs/manuscript/manuscript.tex
Updates from Overleaf

\section{Methods}
%\hl{In this section, describe the data sources and the variables, along with data preparation procedures, variable selection criteria, etc. You should have a table summarizing the water accessibility variables and their definitions. Another table can summarize the explanatory variables (e.g. transportation-related, income-related, etc). Also discuss the number of countries, etc. I've started a table as an example. You can add other columns if needed.}
\subsection{Data Sources}
Data for this study were collected from the Demographic Health Survey (DHS) program, which is a program that has collected, analyzed, and distributed accurate and representative data on population, health, HIV, and nutrition through more than 400 surveys in over 90 countries. DHS the program uses a wide range of monitoring and impact evaluation indicators in the areas of population, health, and nutrition. Also, Standard DHS Surveys have high sample sizes (between 5,000 and 30,000 households) and are administered about every 5 years, to enable comparisons over time. DHS program also uses Malaria Indicator Surveys (MIS) is a survey where they provide data on bed-net ownership and use inhibition of malaria during pregnancy, and timely and adequate treatment of fever in young children. In many cases, bio-marker experiments for malaria and anemia are also included. We used the STAT-Compiler tool available from the DHS Program to develop customized tables on indicators related to water, transportation, and wealth from 78 countries between 1990-2019.

@@

We preserved 28 variables from 78 where less than 50\% of values were missing and the population survey data were removed because they were similar to the household surveys. Ultimately, the simplified data frame was 78 countries by 28 variables. After the simplification, two separate tables are distinguished within the 28 variables. One focused on the water accessibility variables and the other focused on the explanatory variables \citep{price2019difference}. From the remaining variables, we selected 22 which related to water accessibility.  These included: types of water source (improved water source,
piped into dwelling, piped into yard/plot,
 public tap/standpipe, tube well/borehole, protected well, protected spring, rainwater, unimproved water source, unprotected well, unprotected spring, cart with small tank/drum, tanker truck, bottled water, other water source, surface water), time to water source (less than 30 minutes, more than 30 minutes), and premise to water source (water on the premises, improved water source on the premises, basic water service, limited water service) \autoref{tab:summary-wa}. Remaining variables related to wealth and transportation were used as an explanatory variables in the model \autoref{tab:summary-var}.

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{phom} & Piped into dwelling &
 \\\midrule
 \texttt{pipy} & Water piped into yard/plot
&
\\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & 
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pwel} & Protected well
& 
\\\midrule
\texttt{pspr} & Protected spring
& 
\\\midrule

\texttt{rain} & Rainwater
& 
\\\midrule
\texttt{uwel} & Unprotected well water
&
\\\midrule
\texttt{uspr} & Unprotected spring
& 
\\\midrule

 \texttt{truc} & Tanker truck & 
\\\midrule
 \texttt{ctan} & Cart with tank & 
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{othw} & Other water source & 
\\\midrule
\texttt{surw} & Surface water
& 
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
&
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & 
\\\midrule
 \texttt{watp} & Water on the premises
 & 
 \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-wa}
\end{table}

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{bicy} & Bicycle
 &
\\\midrule
 \texttt{acar} & Animal drawn cart
 & 
\\\midrule
 \texttt{mcyc} & Motorcycle
 & 
\\\midrule
 \texttt{pcar} & Private car
 & 
\\\midrule
 \texttt{cgdp} & GDP (current US \$) 
 & [NY.GDP.MKTP.CD]
\\\midrule
 \texttt{tpop} & Total Population
 &  [SP.POP.TOTL]
\\\midrule
 \texttt{upop} & Urban population 
 & (\% of total population)  [SP.URB.TOTL.IN.ZS]
\\\midrule
 \texttt{popd} & Population density 
 & (people per sq. km of land area)  [EN.POP.DNST]
\\\midrule
 \texttt{land} & Land area 
 & (sq. km) [AG.LND.TOTL.K2]
\\\midrule
 \texttt{lita} & Literacy rate, adult total 
 & (\% of people ages 15 and above) [SE.ADT.LITR.ZS]
\\\midrule
 \texttt{lity} & Literacy rate, youth 
 & (ages 15-24), gender parity index (GPI) [SE.ADT.1524.LT.FM.ZS]
\\\midrule
 \texttt{mori} & Mortality rate, under-5 
 & (per 1,000 live births) [SH.DYN.MORT]
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country.

>>>>

\section{Methods}
%\hl{In this section, describe the data sources and the variables, along with data preparation procedures, variable selection criteria, etc. You should have a table summarizing the water accessibility variables and their definitions. Another table can summarize the explanatory variables (e.g. transportation-related, income-related, etc). Also discuss the number of countries, etc. I've started a table as an example. You can add other columns if needed.}
\subsection{Data Sources}
Data for this study were collected from the Demographic Health Survey (DHS) program, which is a program that has collected, analyzed, and distributed accurate and representative data on population, health, HIV, and nutrition through more than 400 surveys in over 90 countries \citep{statcompiler}. DHS the program uses a wide range of monitoring and impact evaluation indicators in the areas of population, health, and nutrition. Also, Standard DHS Surveys have high sample sizes (between 5,000 and 30,000 households) and are administered about every 5 years, to enable comparisons over time. DHS program also uses Malaria Indicator Surveys (MIS) is a survey where they provide data on bed-net ownership and use inhibition of malaria during pregnancy, and timely and adequate treatment of fever in young children. In many cases, bio-marker experiments for malaria and anemia are also included. We used the STAT-Compiler tool available from the DHS Program to develop customized tables on indicators related to water, transportation, and wealth from 73 countries between 2000-2019 \citep{statcompiler}.

We also used World Bank's database to gain additional explanatory variable data for this study. World Bank (WB) provides over 200,000 published documents to understand the development policies and programs globally. Their reports, working papers, documents, and primary data are free to the public. WB DataBank is an analysis and visualization instrument that contains examples of time series data on many topics. We generated tables to analyze them by R.  We used World Development Indicators (WDI) within the DataBank is the primary World Bank collection of development indicators, organized from officially-recognized international sources. It exhibits the most current, accurate global development, and available data includes national, regional, and global estimates. We collected the 2018 surveys for the 73 countries for the following variables mentioned in \autoref{tab:summary-var}.

@@

We preserved 28 variables from 73 countries where less than 50\% of values were missing and the population survey data were removed because they were similar to the household surveys. Ultimately, the simplified data frame was 78 countries by 28 variables. After the simplification, two separate tables are distinguished within the 28 variables. One focused on the water accessibility variables and the other focused on the explanatory variables \citep{price2019difference}. From the remaining variables, we selected 17 which related to water accessibility.  These included: types of water source (piped into dwelling, piped into yard/plot,
 public tap/standpipe, tube well/borehole, protected well, protected spring, rainwater, unprotected well, unprotected spring, cart with small tank/drum, tanker truck, bottled water, other water source, surface water), time to water source (less than 30 minutes, more than 30 minutes), and premise to water source (water on the premises) \autoref{tab:summary-wa}. Remaining variables related to wealth and transportation were used as an explanatory variables in the model \autoref{tab:summary-var}.

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{phom} & Piped into dwelling & DHS Program Surveys
 \\\midrule
 \texttt{pipy} & Water piped into yard/plot
& DHS Program Surveys
\\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & DHS Program Surveys
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pwel} & Protected well
& DHS Program Surveys
\\\midrule
\texttt{pspr} & Protected spring
& DHS Program Surveys
\\\midrule

\texttt{rain} & Rainwater
& DHS Program Surveys
\\\midrule
\texttt{uwel} & Unprotected well water
& DHS Program Surveys
\\\midrule
\texttt{uspr} & Unprotected spring
& DHS Program Surveys
\\\midrule

 \texttt{truc} & Tanker truck & DHS Program Surveys
\\\midrule
 \texttt{ctan} & Cart with tank & DHS Program Surveys
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john (DHS Program Surveys).
\\\midrule
 \texttt{othw} & Other water source & DHS Program Surveys
\\\midrule
\texttt{surw} & Surface water
& DHS Program Surveys
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
& DHS Program Surveys
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & DHS Program Surveys
\\\midrule
 \texttt{watp} & Water on the premises
 & DHS Program Surveys
 \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-wa}
\end{table}

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{bicy} & Bicycle
 & DHS Program Surveys
\\\midrule
 \texttt{acar} & Animal drawn cart
 & DHS Program Surveys
\\\midrule
 \texttt{mcyc} & Motorcycle
 & DHS Program Surveys
\\\midrule
 \texttt{pcar} & Private car
 & DHS Program Surveys
\\\midrule
 \texttt{cgdp} & GDP (current US \$) 
 & World Bank Data [NY.GDP.MKTP.CD]
\\\midrule
 \texttt{tpop} & Total Population
 & World Bank Data   [SP.POP.TOTL]
\\\midrule
 \texttt{upop} & Urban population 
 & World Bank Data  (\% of total population)  [SP.URB.TOTL.IN.ZS]
\\\midrule
 \texttt{popd} & Population density 
 & World Bank Data  (people per sq. km of land area)  [EN.POP.DNST]
\\\midrule
 \texttt{land} & Land area 
 & World Bank Data  (sq. km) [AG.LND.TOTL.K2]
\\\midrule
 \texttt{lita} & Literacy rate, adult total 
 & World Bank Data  (\% of people ages 15 and above) [SE.ADT.LITR.ZS]
\\\midrule
 \texttt{lity} & Literacy rate, youth 
 & World Bank Data  (ages 15-24), gender parity index (GPI) [SE.ADT.1524.LT.FM.ZS]
\\\midrule
 \texttt{mori} & Mortality rate, under-5 
 & World Bank Data  (per 1,000 live births) [SH.DYN.MORT]
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country (World Bank Data).
